---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(ggplot2)
library(Cairo)
```

# tunevt

<!-- badges: start -->
<!-- badges: end -->

The goal of `tunevt` is to implement methods to fit Virtual Twins models (Foster 
et al. (2011)) for identifying subgroups with differential effects in the 
context of clinical trials while controlling the Type I error of falsely 
detecting a differential effect when the conditional average treatment effect is
uniform across the study population.


Virtual Twins is a two-step approach to detecting differential treatment
effects. Subjects' conditional average treatment effects (CATEs) are first
estimated in Step 1 using a flexible model. Then, a simple and interpretable
model is fit in Step 2 to model these estimated CATEs as a function of the
covariates.

The Step 2 model is dependent on some tuning parameter. This parameter is
selected to control the Type I error rate by permuting the data under the
null hypothesis of a constant treatment effect and identifying the minimal
null penalty parameter (MNPP), which is the smallest penalty parameter that
yields a Step 2 model with no covariate effects. The $1-\alpha$ quantile
of the distribution of is then used to fit the Step 2 model on the original
data.

## Installation

`tunevt` is not currently available on [CRAN](https://CRAN.R-project.org).

You can download the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jackmwolf/tunevt")
```
## Example

We consider simulated data from a small clinical trial with 200 subjects.
Each subject has 10 measured covaraites, 8 continuous and 2 binary.

```{r}
library(tunevt)
data("tunevt_example")
head(tunevt_example)
```

We will consider a Virtual Twins model using a random forest to estimate the CATEs in Step 1 and then fitting a regression tree on the estimated CATEs in Step 2 with the Type I error rate set at $\alpha = 0.2$.

```{r cache=TRUE}
set.seed(100)
vtmod <- tunevt(
  data = tunevt_example, Y = "Y", Trt = "Trt", step1 = "randomforest",
  step2 = "rtree", alpha0 = 0.2, p_reps = 100, ntree = 50
)
```

The fit Step 2 model can be accessed via
```{r}
vtmod$vtmod
```

(_We note that the true data generating mechanism_ ($Y_i = h(X_i) + T_i g(X_i)$) _included an interaction between the treatment and whether_ $V_1$ _was above its true mean [sample mean `r round(mean(tunevt_example$V1), 2)`] with_ $g(X_i) = c + 4I(V_{1i}>\mu_1)+4V_{9i}$. 
_So, the procedure did not make a Type I error!_)

We can also look at the null distribution of the MNPP (through `vtmod$theta_null`) and compare `vtmod$mnpp`  $=\widehat\theta$, the MNPP of our data, to  $\theta_\alpha$, the $1-\alpha$ percentile of the null distribution.
Since $\widehat\theta > \theta_\alpha$ we ended up with a Step 2 model that included at least one covariate.

```{r mnpp_plot, dev='CairoPNG', echo = FALSE}
theta_20 <- quantile(vtmod$theta_null, 1 - 0.2, type = 2)

annotate_df <- data.frame(
  x = c(vtmod$mnpp, theta_20),
  label = c("hat(theta)", "theta[alpha]")
)

ggplot(data = NULL, aes(x = vtmod$theta_null)) +
  geom_density(fill = "blue", alpha = 0.1) +
  geom_text(
    data = annotate_df,
    aes(x = x, y = 0, label = label), 
    parse = TRUE, hjust = 0, vjust = 1, nudge_y = Inf, nudge_x = 0.005,
    size = 10
  ) +
  geom_vline(
    aes(xintercept = vtmod$mnpp),
    linetype = 2
    ) +
  geom_vline(
    aes(xintercept = theta_20),
    linetype = 3
  ) +
  labs(
    title = "Null distribution of the Minimal Null Penalty Parameter",
    x = "Regression Tree Complexity Parameter",
    y = "Density"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_classic() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

## References

- Foster, J. C., Taylor, J. M., & Ruberg, S. J. (2011). Subgroup identification from randomized clinical trial data. _Statistics in Medicine, 30_(24), 2867–2880. https://doi.org/10.1002/sim.4322
